A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-01 paper/writeup.tex

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,S]{Peter Ralph}

@@

In the simple example in \figref{fig:conceptual},
extending haplotypes replaces three edges
($0 \to 3$ and $3 \to 4$ on the left tree, and $0 \to 4$ on the right tree)
by two edges ($0 \to 3$ and $3 \to 4$ on both trees).
If all edge endpoints were unique, then we'd expect \emph{every} edge to be extendable
on one of its ends
(except those pendant to the root and some of those adjacent to chromosome ends),
leading to a reduction in number of edges by 1/3.
Experiments with an earlier version of the algorithm showed that
if we only extend haplotypes on ``paths of length 1'' like this, 
then this is achieved for long sequences.
It is possible for \algorithmref{alg:extend} to add edges, as in \figref{extending_diagram},
but we still expect the number of edges to decrease by more than 1/3.
Indeed,
\figref{fig:num_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
in half, as long as the sequence is long enough.

@@

Our next task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
are indeed correct -- i.e., that in addition to compression, we are also gaining information.
To do this, we simulate ARGs containing full haplotypes using \msprime{}
(as described in the Methods).

When we apply our algorithm to the simplified given tree sequence, the ratio of unary span bordering coalescing between the extend tree sequence and the original is 0.4917. This means that \comment{something something something...}  
See \figref{fig:correct-unary-span-log}, \figref{fig:correct-unary-span}, \figref{fig:node-spans}.

@@

@@

The first thing we see is that coalescent nodes are unary
over a substantial portion of their spans:
since the \texttt{simplify} operations removes these unary spans,
the ``simplified'' line on the right of Figure~\ref{fig:discrepancy} shows
the proportion of nodes' spans on which they are not unary.
For instance, on an ARG of 10 samples on a sequence of length $5 \times 10^7$,
only about 15\% of the total spans are retained after simplification;
this rises to around 50\% for 1,000 samples (top right plot)

\begin{figure}[bht]
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
	\caption{
        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
        \comment{TODO: nodes should be in the same order in all three lines,
        using the ordering from the true tree sequence.}
        \comment{TODO: add plot with percent wrongly added by node?}
    }
	\label{fig:node-spans}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Left:} Discrepancy to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (equation~\eqref{eq:discrepancy}).
        \textbf{Right:} Agreement to the true ARG,
        as proportion of the true ARG that is represented.
        The largest number of trees in the bottom row corresponds
        to the sequence length in the top row;
        see text for more details.
    }
    \label{fig:discrepancy}
\end{figure}

\begin{figure}
    \includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_reduction_prop_vs_SL}
    \includegraphics{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_SL}
    \caption{
        \textbf{(A)} Proportion reduction in number of edges
        achieved by extending haplotypes (\algorithmref{alg:extend})
        on tree sequences of various sequence lengths and numbers of samples.
        See Supplementary \figref{sfig:num_edges} for the raw number of edges.
        \textbf{(B)} Runtime for computation of Tajima's $D$
        (using the \tskit method) on the same tree sequences as in (A).
        \comment{TODO: remove shortest lengths and double-check runtime.}
        \label{fig:num_edges}
    }
\end{figure}

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
\end{figure}

@@

\begin{figure}
	\includegraphics{dissimilarity.pdf}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/num_trees_vs_Num_edges_subplot.pdf}
    \caption{
        Raw numbers of edges before and after extending haplotypes
        for the tree sequences shown in \figref{fig:num_edges}
        \label{sfig:num_edges}
    }
\end{figure}

>>>>

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Halley Fritze}
\author[$\dagger$]{Nathaniel Pope}
\author[$\dagger$]{Ava Bamforth}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\ast$,$\dagger$,S]{Peter Ralph}

@@

\comment{TODO: put in benchmarking figures}

In the simple example in \figref{fig:conceptual},
extending haplotypes replaces three edges
($0 \to 3$ and $3 \to 4$ on the left tree, and $0 \to 4$ on the right tree)
by two edges ($0 \to 3$ and $3 \to 4$ on both trees).
If all edge endpoints were unique, then we'd expect \emph{every} edge to be extendable
on one of its ends
(except those pendant to the root and some of those adjacent to chromosome ends),
leading to a reduction in number of edges by 1/3.
Experiments with an earlier version of the algorithm showed that
if we only extend haplotypes on ``paths of length 1'' like this, 
then this is achieved for long sequences.
It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
but we still expect the number of edges to decrease by more than 1/3.
Indeed,
\figref{fig:num_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
in half, as long as the sequence is long enough.

@@

Our next task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
are indeed correct -- i.e., that in addition to compression, we are also gaining information.
To do this, 
we simulate ARGs containing full haplotypes using \msprime{},
apply the simplification algorithm to reduce these so that there are no unary nodes
(i.e., any node present in a marginal tree is a coalescent node or a sample),
and then apply \algorithmref{alg:extend} to the result
(see Methods for more detail).
The method can potentially extend the spans of each node
(spans on which the node will be unary);
and we can quantify how much of these extended spans were in the original ARG
(and thus correctly extended).

\comment{TODO: figure out if we can remove this from figures. If so, move this to Methods.}
Some stuff can't be inferred since it's "isolated unary";
for instance with 1,000 samples, 5e7 that's about half the ARG.
Etcetera.

@@

As seen in \figref{fig:node-spans},
Look, we add lots of stuff
and most most of it is right;
this applies to both recent and ancient nodes;
we're typically extending the span of a node by like maybe 50\%.

\begin{figure}[bht]
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
	\caption{
        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
        \comment{TODO: nodes should be in the same order in all three lines,
        using the ordering from the true tree sequence.}
        \comment{TODO: somehow communicate per-node amount/proportion of correctly and incorrectly added span}
    }
	\label{fig:node-spans}
\end{figure}

\comment{merge this in}
The first thing we see is that coalescent nodes are unary
over a substantial portion of their spans:
since the \texttt{simplify} operations removes these unary spans,
the ``simplified'' line on the right of Figure~\ref{fig:discrepancy} shows
the proportion of nodes' spans on which they are not unary.
For instance, on an ARG of 10 samples on a sequence of length $5 \times 10^7$,
only about 15\% of the total spans are retained after simplification;
this rises to around 50\% for 1,000 samples (top right plot)

We can also use measures of dissimiliarity and similarity to quantify summaries across nodes:
these are shown in \figref{} (purple lines only)
Points:
proportion left after smiplify varies a LOT: from 15\% to 95\%.
    (goes up with L: with longer sequence length, greater opportunity for distant chunks to be on the same lineage)
extending adds hardly any wrong stuff (left plots); and adds a substantial chunk: with large samples, adds like half of the missing stuff

It's also interesting to  note that inference
(see \figref{fig:inferred_edge_counts})
(a) increases number of edges;
(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
(c) but then extending reduces to below the original.

\begin{figure}
	\begin{center}
		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
        \includegraphics[width=\textwidth]{dissimilarity.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Left:} Discrepancy to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (equation~\eqref{eq:discrepancy}).
        \textbf{Right:} Agreement to the true ARG,
        as proportion of the true ARG that is represented.
        The largest number of trees in the bottom row corresponds
        to the sequence length in the top row;
        see text for more details.
        \comment{TODO: make "inferred simplified" green (dotted) not red}
        \comment{TODO: label "inferred extended" "inferred simplfieid extended"??}
        \comment{TODO: also add a "inferred extended" line}
    }
    \label{fig:discrepancy}
\end{figure}

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
    \caption{
        number of edges, 5e7, 1,000 samples
        \label{fig:inferred_edge_counts}
    }
\end{figure}

@@

Main points: in \figref{fig:discrepancy}
(green line) tsinfer gets a lot right (above 60\% with 1000 samples; bottom-right plot)
    and does a lot better with more samples;
    also, gets 10\%--25\% wrong (left panels)
    with longer sequence, larger proportion is wrong (bottom-left panel)
(red dotted)
    simplifying removes a lot of the wrong stuff
    but also a lot of the right stuff
(red solid)
    similarly extending the simplified one adds a bunch of right and also wrong stuff

@@

@@

@@
